---
title: "Project Code"
author: "Minghe Wang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(naniar)
library(knitr)
```


**Research Question:** "Which patient- or hospital-level factors are most predictive of readmission within 30 days among diabetic patients?"

# Data Acquisition and Preprocessing

```{r load_data, message=FALSE}
diabete <- read_csv("./diabetic_data_2.csv")
#head(diabete, 3)
```

About **missingness** in data, three variables(`payer_code`, `medical_spacialty`, `weight`) are dropped due to 10% or more missing values. Although they are direct or indirect related to the readmission outcome, too much missingness is very likely to make them ineffective predictor in the model. Now rows are dropped due to excessive missing values.

```{r clean_data}
# Handle missingness
diabete[diabete == "?"] <- NA
gg_miss_var(diabete)

# Set missingness thresholds
col_na_thresh <- 0.1
row_na_thresh <- 0.1

# Filter out columns with >20% missing values
diabete <- diabete %>%
  dplyr::select(where(~ mean(is.na(.)) <= col_na_thresh))

# Filter out rows with >20% missing values
diabete <- diabete %>%
  filter(rowMeans(is.na(.)) <= row_na_thresh)

# Recode outcome variable
diabete <- diabete %>%
  mutate(readmission_binary = if_else(readmitted == "<30", 1, 0)) %>% 
  mutate(factor(readmission_binary))
```

# Exploratory Data Analysis

```{r EDA}
# Convert Age to ordered factor
age_levels <- c("[0-10)", "[10-20)", "[20-30)", "[30-40)", "[40-50)",
                "[50-60)", "[60-70)", "[70-80)", "[80-90)", "[90-100)")

diabete <- diabete |>
  mutate(age = factor(age, levels = age_levels, ordered = TRUE))

# Summarize key variables
key_vars <- diabete %>% 
  select(race, gender, age, admission_type_id, discharge_disposition_id,
         admission_source_id, A1Cresult, insulin,
         diabetesMed, readmission_binary, diag_1, time_in_hospital)
summary(key_vars)

# Format reader freindly frequency table
key_vars_list <- c("race", "gender", "age", "admission_type_id", "discharge_disposition_id",
              "admission_source_id", "A1Cresult",
              "insulin", "diabetesMed", "readmission_binary", "time_in_hospital")

plots <- list()

for (var in key_vars_list) {
  cat("##", var)
  freq_table <- diabete |>
    count(across(all_of(var)), name = "Count") |>
    mutate(Percent = round(Count / sum(Count) * 100, 1)) |>
    arrange(desc(Count)) |>
    rename(Level = 1)
  
  print(kable(freq_table))
  cat("\n\n")
  
  # Bar plots
  plot <- ggplot(freq_table, aes(x = fct_infreq(as.factor(Level)), y = Count)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    labs(title = var, x = "Category", y = "Count") +
    theme_minimal(base_size = 8) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1)) + 
    theme(plot.title = element_text(hjust = 0.5))
  
  plots[[var]] <- plot
}
#plots
gridExtra::grid.arrange(grobs = plots, nrow = 3, ncol = 4)
```


# Model Development

# Posterior Inference and Model evaluatin

# Interpretation and Reporting


